---
title: "Reproducible Research Course Project 1"
author: "Enrique Estrada"
---

## Loading and preprocessing the data
  Unzip and load data into data frame `data`.
```{r}
if(!file.exists('activity.csv')){
  unzip('activity.zip')
}
data <- read.csv("activity.csv")
```

## 1. What is mean total number of steps taken per day?
```{r echo=TRUE}
steps_by_day <- aggregate(steps ~ date, data, sum)
```
* Make a histogram of the total number of steps taken each day
```{r echo=TRUE}
hist(steps_by_day$steps, main = "Total number of steps taken per day", xlab = "Total steps taken per day", col = "darkblue", ylim = c(0,20), breaks = seq(0,25000, by=2500))

```

* Calculate and report the **mean** and **median** total number of steps taken 
per day
```{r echo=TRUE}
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```
The `mean` is `r rmean` and the `median` is `r rmedian`.

## 2. What is the average daily activity pattern?
* Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r echo=TRUE}
steps_by_interval <- aggregate(steps ~ interval, data, mean)
plot(steps_by_interval$interval,steps_by_interval$steps, type="l", col="darkblue", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")
```

* Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps? 
```{r echo=TRUE}
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```
Most Steps at:`r max_interval`.

## 3. Imputing missing values 
* Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r echo=TRUE}
incomplete <- sum(is.na(data))
```
Number of missing values:`r incomplete`. 

* Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Compare imputed to non-imputed data. Missing values were imputed by inserting the average for each 
interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days
(0.1320755), replaced the NA. 
```{r echo=TRUE}
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
```

* Create a new dataset that is equal to the original dataset but with the missing data filled in.

Zeroes were imputed for 10-01-2012 because it was the first day and would have been over 9,000 steps 
higher than the following day, which had only 126 steps. NAs then were assumed to be zeros to fit the rising trend of the data. 
```{r echo=TRUE}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

* Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day.

Recount total steps by day and create Histogram. 
```{r echo=TRUE}
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference.

hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T) 
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Calculate new mean and median for imputed data. 
```{r, echo=TRUE}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

Calculate difference between imputed and non-imputed data.
```{r, echo=TRUE}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```

Calculate total difference.
```{r, echo=TRUE}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```
* The imputed data mean is `r rmean.i`
* The imputed data median is `r rmedian.i`
* The difference between the non-imputed mean and imputed mean is `r mean_diff`
* The difference between the non-imputed mean and imputed mean is `r med_diff`
* The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Thus, there were `r total_diff` more steps in the imputed data.


## 4. Are there differences in activity patterns between weekdays and weekends?
* Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
# For Sunday and Saturday : weekend, Other days : weekday 
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))
```

* Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data
```{r, echo=TRUE}
steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)
library(lattice)
xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```

